Type Classes for Mathematics in Type Theory

### 2010-07-01 jar.tex
Sacrifice !'s in parameter declarations to improve presentation.

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
      `{!CatId O} `{!CatComp O}: Prop :=
    { arrow_equiv:> Π x y, Setoid (x ⟶ y)
    ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: Π s, eqt_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
\end{lstlisting}
  % todo: why do i need the manual \hspace!?

@@

For the homomorphism theorem, we begin by declaring our dramatis personae:
\begin{lstlisting}
  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{!HomoMorphism σ$$ A B f}.
\end{lstlisting}
With \lstinline|~| defined as described, the first part of the proof is simply the definition of the following instance:
\begin{lstlisting}
  Instance co: Congruence σ$$ (~).
\end{lstlisting}

@@

\lstinline|EquationalTheory|'s for semirings and rings are easy to define, and so from section~\ref{varieties} we get corresponding categories in which we can postulate initial objects:
\begin{lstlisting}
  Class Naturals (A: ObjectInVariety semiring_theory) `{!InitialArrow A}: Prop :=
    { naturals_initial:> Initial A }.
\end{lstlisting}
While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component `leaks' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice. Furthermore, this definition needs an additional layer of class instances to relate it to the \lstinline|SemiRing| class from our algebraic hierarchy.

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{!NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> Π `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, and arrays of native machine words.

@@

The \lstinline|Quote| instance for multiplication is a little more subtle, but really only does a bit of heap juggling:
\begin{lstlisting}
  Instance quote_mult V (v: Env V) n V' (v': Env V') m V'' (v'': Env V'')
    `{!Quote v n v'} `{!Quote (merge v v') m v''}:
      Quote v (n * m) (merge v' v'') :=
      { quote :=
        Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
\end{lstlisting}

@@

  Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
    := { key := inl (key x va) }.

  Instance lookup_right `{!Lookup x vb}: Lookup x (merge va vb)
    := { key := inr (key x vb) }.
\end{lstlisting}

@@

The following additional utility lemma lets us quote equalities with a shared heap (so that an opaque expression that occurs on both sides of the equation is not represented by two distinct variables):
\begin{lstlisting}
  Lemma quote_equality {V} {v: Env V} {V'} {v': Env V'} (l r: Value)
    `{!Quote novars l v} `{!Quote v r v'}:
      let heap := merge v v' in
      eval heap (map_var shift quote) = eval heap quote → l = r.
\end{lstlisting}

>>>>

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> Π x y, Setoid (x ⟶ y)
    ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: Π a, Equiv (carriers a)} `{AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: Π s, eqt_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
\end{lstlisting}
  % todo: why do i need the manual \hspace!?

@@

For the homomorphism theorem, we begin by declaring our dramatis personae:
\begin{lstlisting}
  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{HomoMorphism σ$$ A B f}.
\end{lstlisting}
With \lstinline|~| defined as described, the first part of the proof is simply the definition of the following instance:
\begin{lstlisting}
  Instance co: Congruence σ$$ (~).
\end{lstlisting}

@@

\lstinline|EquationalTheory|'s for semirings and rings are easy to define, and so from section~\ref{varieties} we get corresponding categories in which we can postulate initial objects:
\begin{lstlisting}
  Class Naturals (A: ObjectInVariety semiring_theory) `{InitialArrow A}: Prop :=
    { naturals_initial:> Initial A }.
\end{lstlisting}
While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component `leaks' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice. Furthermore, this definition needs an additional layer of class instances to relate it to the \lstinline|SemiRing| class from our algebraic hierarchy.

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> Π `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, and arrays of native machine words.

@@

The \lstinline|Quote| instance for multiplication is a little more subtle, but really only does a bit of heap juggling:
\begin{lstlisting}
  Instance quote_mult V (v: Env V) n V' (v': Env V') m V'' (v'': Env V'')
    `{Quote v n v'} `{Quote (merge v v') m v''}:
      Quote v (n * m) (merge v' v'') :=
      { quote :=
        Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
\end{lstlisting}

@@

  Instance lookup_left `{Lookup x va}: Lookup x (merge va vb)
    := { key := inl (key x va) }.

  Instance lookup_right `{Lookup x vb}: Lookup x (merge va vb)
    := { key := inr (key x vb) }.
\end{lstlisting}

@@

The following additional utility lemma lets us quote equalities with a shared heap (so that an opaque expression that occurs on both sides of the equation is not represented by two distinct variables):
\begin{lstlisting}
  Lemma quote_equality {V} {v: Env V} {V'} {v': Env V'} (l r: Value)
    `{Quote novars l v} `{Quote v r v'}:
      let heap := merge v v' in
      eval heap (map_var shift quote) = eval heap quote → l = r.
\end{lstlisting}
